Awareness and Russell's Epistemology

### 2011-01-23 awareness.tex
Lots of reorganizing the intro

In this section, we give the explicit semantics to basic awareness logic,
originally presented in {\cite{fagin_belief_1988}}, with an additional axiom.
 One difference in our presentation is the inclusion of a novel letter and a
superficial change in notation.  This work is inspired by the developments in
van Benthem and Vel\'azquez-Quesada {\cite{benthem_inference_2009}}.

@@

Intuitively, the \tmtextbf{CSQ} asserts that an agent believes a formula
$\phi$ if and only if it follows logically from their knowledge base and their
background knowledge, represented by $\tmop{Th} (\mathbbm{M})$. 
\tmtextbf{CSQ} is Vincent Hendrik's principle of logical omniscience mentioned
in {\S}\ref{intro}.

\tmtextbf{SND} asserts that $\circlearrowleft$ corresponds to the agent's
knowledge base being sound. A sound knowledge base will only render true
conclusions when used in deductions.  If one considers ``the existence of a
sound deduction'' sufficient for a kind of knowledge, then $\circlearrowleft$
is suitable for investigating this notion of epistemology.  That is, if
$\mathbbm{M}, w \vDash \circlearrowleft \wedge \Box \phi$, then we may
naturally say that the agent (implicitly) {\tmem{knows}} $\phi$ on the basis
of $\mathcal{A}_w$, rather than merely {\tmem{believing}} $\phi$ on the basis
of $\mathcal{A}_w$.

>>>>

In this section, we give the explicit semantics to basic awareness logic,
originally presented in {\cite{fagin_belief_1988}}, with an additional axiom.
%  One difference in our presentation is the inclusion of a novel letter and a
% superficial change in notation. 
 This work was inspired by the developments in
van Benthem and Vel\'azquez-Quesada {\cite{benthem_inference_2009}}.

@@

### 2011-01-23 awareness0.tex
Lots of reorganizing the intro

In this section, we give the explicit semantics to basic awareness logic,
originally presented in {\cite{fagin_belief_1988}}, with an additional axiom.
%  One difference in our presentation is the inclusion of a novel letter and a
% superficial change in notation. 
 This work was inspired by the developments in
van Benthem and Vel\'azquez-Quesada {\cite{benthem_inference_2009}}.

\begin{definition}
  Let $\Phi$ be a set of letters and define the language $\mathcal{L}_A
  (\Phi)$ as:
  \[ \phi \hspace{1em} : : = \hspace{1em} p \in \Phi \hspace{1em} |
     \hspace{1em} \circlearrowleft \hspace{1em} | \hspace{1em} \bot
     \hspace{1em} | \hspace{1em} \phi \rightarrow \psi \hspace{1em} |
     \hspace{1em} \Box \phi \hspace{1em} | \hspace{1em} A : \phi \hspace{1em}
  \]
\end{definition}

@@

\begin{definition}
  \label{awarenessmodels}An {\tmstrong{awareness model}} $\mathbbm{M}= \langle
  W, V, R, \mathcal{A} \rangle$ is a Kripke model with a valuation $V : \Phi
  \cup \{\circlearrowleft\} \rightarrow 2^W$ and an awareness function
  $\mathcal{A} : W \rightarrow 2^{\mathcal{L}_A (\Phi)}$.  We will usually
  denote $\mathcal{A} (w)$ as $\mathcal{A}_w$.

  Define the semantic entailment relation $\vDash$ where atomic propositions,
  logical connectives and modality are interpreted as usual.  We employ the
  following novel definition:
  \begin{eqnarray*}
    \mathbbm{M}, w \vDash A : \phi & \text{iff} & \phi \in \mathcal{A_w}
  \end{eqnarray*}
\end{definition}

The following definition distinguishes us from previously investigations of
awareness logic:

\begin{definition}
  Define the ``\tmtextbf{theory of a model} $\mathbbm{M}$'' as:
  \[ \tmop{Th} (\mathbbm{M}) \assign \{\phi \hspace{1em} | \hspace{1em}
     \mathbbm{M}, v \vDash \phi \text{ for all } v \in W\} \]
  We define the following properties that a model $\mathbbm{M}$ may posses:

  \begin{descriptiondash}
    \item[CSQ] $\mathbbm{M}, w \vDash \Box \phi \text{ iff } \tmop{Th}
    (\mathbbm{M}) \cup \mathcal{A}_w \vdash \phi$

    \item[SND] $\mathbbm{M}, w \vDash \circlearrowleft \text{ iff }
    \mathbbm{M}, w \vDash \phi$ for all $\phi \in \mathcal{A}_w$
  \end{descriptiondash}

@@

Neither \tmtextbf{CSQ} nor \tmtextbf{SND} correspond to simple modally
definable properties.

The logic of models making true \tmtextbf{CSQ} and \tmtextbf{SND} is provided
in Table \ref{logic1}.

\begin{table}[h]
  \begin{tabular}{l}
    $\vdash \phi \rightarrow \psi \rightarrow \phi$\\
    $\vdash (\phi \rightarrow \psi \rightarrow \chi) \rightarrow (\phi
    \rightarrow \psi) \rightarrow \phi \rightarrow \chi$\\
    $\vdash ((\phi \rightarrow \bot) \rightarrow (\psi \rightarrow \bot))
    \rightarrow \psi \rightarrow \phi$\\
    $\vdash \Box(\phi \rightarrow \psi) \rightarrow \Box \phi \rightarrow \Box
    \psi$\\
    $\vdash A : \phi \rightarrow \Box \phi$\\
    $\vdash \circlearrowleft \rightarrow \Box \phi \rightarrow \phi$\\
    \\
    \begin{tabular}{lll}
      $\frac{\vdash \phi \rightarrow \psi \hspace{4em} \vdash \phi}{\vdash
      \psi}$ & {\hspace{6em}} & $\frac{\vdash \phi}{\vdash \Box \phi}$
    \end{tabular}
  \end{tabular}
  \caption{\label{logic1}Awareness Logic for \tmtextbf{CSQ} and
  \tmtextbf{SND}}
\end{table}

\begin{theorem}
  \label{completeness1}Assuming an infinite store of proposition letters
  $\Phi$, the awareness logic presented is sound and weakly complete for
  awareness models making true \tmtextbf{CSQ} and \tmtextbf{SND}.
\end{theorem}

>>>>

In this section, apply basic awareness logic to Russell's
epistemology.  As we described in \S\ref{mod-russell}, we will use
awareness to model \emph{acquaintance}, and modality to model \emph{knowledge by description}.

\begin{definition}
  Let $\Phi$ be a set of letters and define the language $\mathcal{L}_A
  (\Phi)$ as:
  \[ \phi \  : : = \  p \in \Phi \  |
     \  \circlearrowleft \  | \  \bot
     \  | \  \phi \rightarrow \psi \  |
     \  \Box \phi \  | \  A : \phi \ 
  \]
\end{definition}

@@

\begin{definition}
  \label{awarenessmodels0}An {\tmstrong{awareness model}} $\mathbbm{M}= \langle
  W, V, R, \mathcal{A} \rangle$ is a Kripke model with a valuation $V : \Phi
  \cup \{\circlearrowleft\} \rightarrow 2^W$ and an awareness function
  $\mathcal{A} : W \rightarrow 2^{\mathcal{L}_A (\Phi)}$.  We will usually
  denote $\mathcal{A} (w)$ as $\mathcal{A}_w$.

  Define the semantic entailment relation $\vDash$ according to
  \textbf{awareness logic} as defined in \cite{fagin_belief_1988},
  where $A : \phi$ denotes awareness of $\phi$. 
% where atomic propositions,
%   logical connectives and modality are interpreted as usual.  We employ the
%   following novel definition:
%   \begin{eqnarray*}
%     \mathbbm{M}, w \vDash A : \phi & \text{iff} & \phi \in \mathcal{A_w}
%   \end{eqnarray*}
\end{definition}

Define the following property, as sketched in \S\ref{mod-russell}:

\begin{definition}
  \begin{descriptiondash}
    \item[CSQ$_\vdash$] $\mathbbm{M}, w \vDash \Box \phi \text{ iff } \tmop{Th}
    (\mathbbm{M}) \cup \mathcal{A}_w \vdash \phi$
    % \item[SND] $\mathbbm{M}, w \vDash \circlearrowleft \text{ iff }
    % \mathbbm{M}, w \vDash \phi$ for all $\phi \in \mathcal{A}_w$
  \end{descriptiondash}

@@

% Neither \tmtextbf{CSQ} nor \tmtextbf{SND} correspond to simple modally
% definable properties.

The logic of models making true \tmtextbf{CSQ} % and \tmtextbf{SND} 
is provided in Table \ref{logic0}.

\begin{table}[h]
  \begin{tabular}{l}
    $\vdash \phi \rightarrow \psi \rightarrow \phi$\\
    $\vdash (\phi \rightarrow \psi \rightarrow \chi) \rightarrow (\phi
    \rightarrow \psi) \rightarrow \phi \rightarrow \chi$\\
    $\vdash ((\phi \rightarrow \bot) \rightarrow (\psi \rightarrow \bot))
    \rightarrow \psi \rightarrow \phi$\\
    $\vdash \Box(\phi \rightarrow \psi) \rightarrow \Box \phi \rightarrow \Box
    \psi$\\
    $\vdash A : \phi \rightarrow \Box \phi$\\
    $\vdash \circlearrowleft \rightarrow \Box \phi \rightarrow \phi$\\
    \\
    \begin{tabular}{lll}
      $\frac{\vdash \phi \rightarrow \psi \hspace{4em} \vdash \phi}{\vdash
      \psi}$ & {\hspace{6em}} & $\frac{\vdash \phi}{\vdash \Box \phi}$
    \end{tabular}
  \end{tabular}
  \caption{\label{logic0}Awareness Logic for Acquaintance and
    Knowledge by Description}
\end{table}

\begin{theorem}
  \label{completeness0}Assuming an infinite store of proposition letters
  $\Phi$, the awareness logic presented is sound and weakly complete for
  awareness models making true \tmtextbf{CSQ} and \tmtextbf{SND}.
\end{theorem}

### 2011-01-23 introduction.tex
Fixed another broken citation
Lots of reorganizing the intro

The modern tradition in epistemic logic is to assume knowledge modalities
conform to the $S 5$ axiom schema.  As noted in
{\cite{halpern_set-theoretic_1999,rubinstein_modeling_1998}}, the semantics of
$S 5$ knowledge correspond exactly to partitioning a collection of situations
into \tmtextit{information sets}, which is the traditional approach in game
theory and decision theory.  While it is not commonly acknowledged in
epistemic logic, economists and philosophers accept that traditional decision
theory is externalist and behaviorist in nature
\footnote{An early essay by
Amartya Sen on the philosophical foundations of traditional decision theory
makes the behaviorist reading of decision theory clear
\cite{sen_behaviour_1973}.  Kaushik Basu also discusses the behaviorist
nature of decision theory {\cite[pgs. 53--54]{basu_revealed_1980}}. 
Finally, Donald Davidson appeals to decision theory to motivate an externalist
epistemology in \cite{davidson_could_1995}.}.

In {\cite{benthem_reflections_1991}}, Johan van Benthem proposed
a research program to find logics for explicit knowledge, providing the first
suggestion of an internalist perspective on epistemic logic.  Subsequently,
Sergei Artemov and Elena Nogina proposed a logic of explicit justification,
which has come to be known as \tmtextit{Justification Logic} (JL)
{\cite{artemov_introducing_2005}}, based on Artemov's \tmtextit{Logic of
Proofs} {\cite{artemov_logic_1994}}.  While the original semantics of
JL were based on interpretability into Peano Arithmetic, Melvin Fitting proposed Kripke
semantics for JL in {\cite{fitting_logic_2005}}.  Recently, Fernando
Vel\'azquez-Quesada and Johan van Benthem have developed a simpler framework
for explicit epistemics in {\cite{benthem_inference_2009}}, entitled the
\tmtextit{Dynamics of Awareness}.  This work was based on Joseph Halpern and
Ronald Fagin's original \tmtextit{Awareness Logic} {\cite{fagin_belief_1987}}.

In this essay we repurpose various externalist logics to take on an
internalist reading.  The concept of a \tmtextit{knowledge base}, from which
beliefs may be implicitly deduced, will play a crucial role in our discussion.
\ We propose this as an avenue for representing foundationalist perspectives
on epistemology in epistemic logic.  Our philosophical motivation is taken
from two sources.  The first is Vincent Hendricks in
{\cite{hendricks_mainstream_2006}}, where he characterizes the principal of
{\tmem{logical omniscience}} for implicit knowledge in epistemic
logic{\footnote{We have modified Hendricks' notation here slightly to match
our own.}}:

\begin{quote}
  {\tmem{Whenever an agent $\Xi$ knows all of the formulae in $\mathcal{A}$,
  and $\phi$ follows logically from $\mathcal{A}$, then $\Xi$ [implicitly]
  knows $\phi$.}}
\end{quote}

We will design our semantics such that ``$\Box \phi$'' may be equated with
``$\phi$ follows logically from a knowledge base $\mathcal{A}$,'' which is
sometimes written as $\phi \in \tmop{Cn} ( \mathcal{A})$ in the artificial
intelligence literature.  Our second inspiration comes from what Hilary
Kornblith calls the \tmtextit{the arguments on paper thesis}
{\cite{kornblith_beyond_1980}}, which he feels characterizes apsychological,
internalist theories of knowledge:

\begin{quote}
  Let us suppose that, for any person, it is possible, at least in principle,
  to list all of the propositions that person believes.  The
  arguments-on-paper thesis is just the view that a person has a justified
  belief that a particular proposition is true just in case that proposition
  appears on the list of propositions that person believes, and either it
  requires no argument, or a {\tmem{good argument}} can be given for it which
  takes as premises certain other propositions on the list.
\end{quote}

Kornblith asserts that foundationalism and coherentism vie for accounts of a
``good argument'' in the above thesis.  He provides an extensive bibliography
citing proposals for this principle by a number of 20th century
epistemologists, including figures such as A.J. Ayer and C.I. Lewis.  The
rest of Kornblith's paper is devoted to attacking this view and proposing a
form of naturalized epistemology; we will not address this debate here,
however.

We adopt a arguments-on-paper-thesis perspective on epistemic logic in this
paper.  We consider a \tmtextit{good argument} to be a logical derivation
from propositions present in a knowledge base in this setting.  \
Special attention will be given to \tmtextit{sound derivations}, which will be
thought of as a form of knowledge.  In \emph{Awareness Logic}, we will interpret
awareness of a formula as membership in a knowledge base.  We prove
completeness for basic awareness logic and a hybrid logic extension.  Logics
of multiple knowledge bases are also presented: a simplified form of
Justification Logic, a logic with neighborhood semantics, an a logic with
modalities for quantifying over knowledge bases.  We conclude with an
application to naturalized epistemology found in the psychological
literature.
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "paper"
%%% End:

>>>>

Modern epistemic logic is commonly acknowledged to begin with
Hintikka's \emph{Knowledge and Belief}
\cite{hintikka_knowledge_1969}. While Hintikka was steeped in
profound philosophical insights, the introduction of epistemic logic
bifurcated 20th century research into epistemology.  A number of
logicians and philosophers have acknowledged this rift, most notably
Vincent Hendricks \cite{hendricks_wheres_2006,
  hendricks_mainstream_2006}.

The purposes of this essay is to try to close the gap between these
traditions, by showing that awareness logics are suitable for
reasoning about the form of foundationalism proposed
by Bertrand Russell in \emph{The Problems of Philosophy}
\cite{russell_problems_1936}.  We begin by reviewing Russell's
theory, and contrasting it with modern perspectives.  In particular,
we illustrate that the modern concepts of \emph{explicit} and
\emph{implicit}, due to Levesque\footnote{While we will refer often to
  the terms coined by Levesque in this essay, we will not use his
  proposed semantics.  We will instead prefer variations on
  \emph{awareness logic}, originally developed by Fagin and Halpern \cite{fagin_belief_1988}.}\cite{levesque_logic_1984}, knowledge in epistemic logic are reinventions of
Russell's distinction between \emph{immediate} and \emph{derivative}
knowledge.  We propose a novel property of awareness models which
directly reflects Russell's foundationalist epistemology.  The
remainder of this paper is then devoted to exploring how awareness
logics can be designed to express our this foundationalist property.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "paper"
%%% End:

### 2011-01-23 paper.tex
Lots of reorganizing the intro

\title{Epistemic Logics For Knowledge Bases}\author{Matthew P.
Wampler-Doty}\date{}\maketitle

\begin{abstract}
  In {\cite{benthem_reflections_1991}}, Johan van Benthem
  proposed a research program for logic of \tmtextit{explicit knowledge}. 
  Two efforts have since emerged with proposals for logic of explicit
  knowledge: the \tmtextit{Dynamics of Awareness}
  {\cite{benthem_inference_2009}} and \tmtextit{Justification Logic}
  {\cite{artemov_introducing_2005}}.  The purpose of this paper is to
  repurpose various epistemic logics from these previous efforts for reasoning
  about logical closures over knowledge bases.  The results in this paper are
  based on previous work carried out in
  {\cite{wampler-doty_evidentialist_2010}}.
\end{abstract}

@@

\section{Basic Explicit Logics of Knowledge Bases}

\subsection{Awareness Logic}
\label{awarenesslogic}
\input{awareness.tex}

\subsection{Hybrid Awareness Logic}
\label{hybrid_awareness}
\input{hybrid_awareness.tex}

\section{Logics of Multiple Knowledge Bases}

So far, we have only dealt with systems that reason over a single knowledge
base.  In this section, we present various logics for reasoning over multiple
knowledge bases.

@@

\subsection{Lattice Justification Logic}
\label{lattice_justification}
\input{lattice_justification.tex}

@@

\subsection{\label{quantifying}Using Modalities to Quantify over Knowledge
Bases}
\label{modal-modal}
\input{modal-modal.tex}

>>>>

\title{Awareness and Russell's Epistemology}\author{Matthew P.
Wampler-Doty}\date{}\maketitle

\begin{abstract}
  In this essay, we show that awareness logic is suitable for
  formal modeling of Bertrand Russell's theory of knowledge,
  originally developed in \cite{russell_problems_1936}.  This is done
  by introducing a novel property to the semantics of 
  awareness logics. 
  We present a variety of sound and complete awareness
  logics compatible which exhibit this property.
\end{abstract}
\todo{This abstract isn't flowery enough.  Lookup PhD comics
  ``Abstract madlibs'' and use that as a template.}

@@

\section{Philosophical Background}
\label{trad}
\input{foundationalism.tex}

\section{Basic Explicit Logics of Foundationalism}

\subsection{Acquaintance and Knowledge By Description}
\label{awarenesslogic0}
\input{awareness0.tex}

\subsection{Knowledge and Belief}
\label{awarenesslogic}
\input{awareness.tex}

\subsection{Hybrid Logic}
\label{hybrid_awareness}
\input{hybrid_awareness.tex}

\section{Logics of Multiple Awareness Sets}

So far, we have only dealt with systems for reasoning over a single knowledge
base.  In this section, we present various logics for reasoning over multiple
knowledge bases.

@@

% \subsection{Lattice Justification Logic}
% \label{lattice_justification}
% \input{lattice_justification.tex}

@@

\subsection{\label{quantifying}Multi-Modal Logic}
\label{modal-modal}
\input{modal-modal.tex}
